Title Pending

### 2019-02-18 manuscript/manuscript.tex
next round of edits ready for Nathaniel

\keywords{
decision theory; computational psychiatry; anxiety; value iteration
}

@@

Central to the experience and maintenance of anxiety disorders are the generalization of fear and persistent avoidance. In fear generalization, the set of perceived threats expands to include secondary cues or situations associated with the primary danger. For example, a dog phobic individual may first develop a fear of dog parks, then parks in general, and finally any public space where a dog may be plausibly encountered. As a consequence, avoidance behavior becomes more prolific often at considerable cost to the anxious individual, either in foregone rewards or in effort expenditure. The same dog phobic individual may take lengthy detours to avoid proximity to feared locations and may even abandon social obligations for fear of encountering a dog. Importantly, anxious avoidance is persistent despite the many safety precautions often in place that limit or negate the possibility of danger (e.g. fences, leash laws). These symptoms are puzzling then insofar the objective risk of danger is often so low and the cost of anxious behavior so high (as is often acknowledged by the patients themselves).

From the perspective of decision theory, present choice (e.g. to approach or avoid) is contingent on assumptions about the future, such as the reliability of the environment or the set of available future choices. A common finding in clinical research is that anxiety is correlated with pessimistic expectations about the future, such that anxious individuals are more likely to endorse a belief in a lack of control over future danger. As has been noted elsewhere \cite{Krypotos2015}, a great many studies of anxious decision making have focused on avoidance behavior in response to immediate danger rather than on early avoidance of potential future threat. The latter are arguably more important to the understanding and treatment of anxiety, as much of its disruption to everyday functioning stems from repeated avoidance of anticipated threat.

Here, we explore these issues by offering a decision theoretic account of sequential choice under pessimistic future expectations. We consider optimal behavior in deterministic, finite Markov decision processes (MDP), in which agents learn reward-maximizing (or loss-minimizing) policies in grid-world environments requiring series of choices. Specifically, our model treats pessimistic expectations as the aberrant assumption that an agent will be unable to act consistently with its preferences in the future. In other words, our model assumes that agents are reward-maximizing manner in present choice but expect that such choices will be unavailable to them in the future. We show that this minor deviation from standard temporal difference learning models captures many features of anxious behavior, including exaggerated threat appraisal, fear generalization, and persistent avoidance.

In the remaining sections, we also demonstrate that this simple model can account for the findings reported in the handful of empirical studies measuring anxious choice in sequential choice environments. We then offer a speculative prediction from the model that, consistent with negative future expectations, anxious individuals should not exhibit the free choice preference observed in non-anxious individuals. Finally, we discuss the relationship of the present model to clinical theories postulating a role for control and helplessness in anxiety disorders.

@@

Crucially then, the value of a state under the optimal policy is contingent on an agent's particular expectations about the future. First, it is a function of the state transition distribution, or beliefs about the likelihood of encountering particular states conditioned on undertaking some action. Second, it is also a function of the actions available in those successor states, as those in turn define the value of these states. Under Bellman optimality, it is normative for an agent to assume that the value of a successor state is equivalent to the best action from that state. This can be expressed as:

@@

where $q(s',a')$ denotes the a state-action value. As we will demonstrate, this assumption has profound consequences for learning about approach and avoidance behaviors.

In a MDP, the discovery of a rewarding action by an agent has an effect not only on the associated state but also on its antecedent states. This follows naturally from the recursive definition of state values defined above. So when an agent learns a particular state predicts reward, the states which immediately precede it also acquire value (as do their antecedent states, and so on) thereby propagating positive value across the state space so as to reinforce reward-maximizing policies. This process does not occur, however, upon the discovery of a suboptimal or aversive action. Because under Bellman optimality the value of a state is dependent in part on the best possible future action, the discovery of harmful actions will not necessarily impact state value estimates or the optimal policy. Insofar that suboptimal or aversive actions can be avoided, then Bellman optimality predicts a fundamental asymmetry: reward opportunities propagate recursively to antecedent states, but danger does not.

This principle is highlighted in Figure 1. In an open field environment, an agent is free to navigate a discrete, deterministic grid-world devoid of rewards except for two states: one rewarding (blue) and one harmful (red, Figure 1a). In this environment, the one-step transition distribution respects the agent's choice; that is, the agent comes into experiences an aversive outcome only if it decides to approach danger. As can be observed, the state value estimates learned by an agent under Bellman optimality (Figure 1b) reflect the asymmetry described above. Despite the presence of danger, the agent learns to associate all states (except the harmful state itself) with positive value as the reward opportunity propagates outwards to all antecedent states. Insofar that danger can be successfully avoided, all states (even those adjacent to the danger) come to predict expected future reward.

What happens, however, when an agent believes that danger is not perfectly avoidable and safety not guaranteed? Such a belief may manifest as inaccurate assumptions about the state transition distribution. For example, an agent may wrongly believe it possible that it could transition to a state proximal to danger in spite of its choices. Alternately, this belief may manifest as pessimistic expectations about the availability of future actions. An agent may believe it has the capacity to act in accordance with the (reward-maximizing) optimal policy in the present, but may be unable to do so in the future. Importantly, such beliefs about the unreliability of the environment or of the self are commonly endorsed in anxiety disorders.

We can explicitly model this type of pessimism about future expectations through a minor change to the state value function defined above. Here, we focus on one particular formulation, previously introduced by \cite{Gaskett2003}, that captures the expectation that future action may deviate from the optimal reward-maximizing policy. We define the value of the successor state as:

@@

where the parameter $w$ controls the degree of pessimism. When $\w = 1$, an agent is maximally optimistic and fully expects to act according to the best possible actions in the future. When $w = 0$, the agent is maximally pessimistic and fully expects to act according to the worst possible actions in the future. When $w = 0.5$, the weighs the best and worst possible future actions equally. We note we are not committed to this particular implementation as anything more than a computational convenience. Many alternate functions could be defined over the space of possible future actions. (As mentioned above, we could also modify the transition distribution.) This formalism highlights an important computational principle; it does not constitute a mechanistic claim about anxiety disorders.

@@

The consequences of this manipulation for the open field environment are presented in Figure 1c,d. When Bellman optimality is violated (i.e. an agent no longer assumes the value of a successor state is equal the best available action), more than reward opportunities are able to propagate value through the state space. Indeed when $w$ is non-zero, the agent treats the future potential of danger as possible and thus negative value is able to spread from its source recursively through antecedent states thereby degrading the value of distal states (the degree of which determined by the extent of pessimism, $w$, and temporal discounting, $\gamma$). With increasing certainty that the worst policy will be followed in the future, the larger the extent to which threat value can permeate.

In addition to drawing inspiration from clinical theories of anxiety, this pessimistic model also captures several core symptoms of anxiety disorders. The hallmark symptom of anxiety disorders is exaggerated threat appraisal, or the tendency to perceive threat as disproportionately greater in likelihood and severity than is justified. This is evident in the anxious state value estimates presented in Figure 1c,d. Indeed when safety is perceived to be uncertain, threat in the environment is exaggerated in magnitude and prevalence. As should be apparent, this very same mechanism captures the generalization of fear that often accompanies anxiety disorders.

In addition to pessimistic inference, all anxiety disorders share pathological avoidance as core symptom \citep{dsm5, Krypotos2015, Arnaudova2017}. Avoidance behaviors seek to increase or maintain the temporal, physical and/or mental distance between an individual and a threatening situation. Although avoidance is a natural response to danger, excessive avoidance as is observed in anxiety disorders can be highly disruptive to everyday functioning \citep{Salter2004} and frequently function so as maintain anxiety by preventing extinction \citep{Arnaudova2017}. For example, an individual with social anxiety may avoid attending large social events for fear of public embarrassment only to never learn that such worry is not justified.

Avoidance behaviors in anxiety have been studied using a variety of methods. Perhaps the most classic paradigm is the behavioral avoidance test \citep{bandura1977}, wherein participants complete a number of tasks involving a feared object or situation and the number of tasks completed (or lack thereof) is a measure of avoidance. Computerized avoidance tasks have become increasingly common, and increased distance from virtual threat has been found to correlate with anxiety \citep{Bach2014, Bach2017, Sheynin2014}. This is not the case for the pessimistic agent. As above, when the agent doubts its future ability to act in a reward-maximizing (loss-minimizing) manor, it cannot rule out the possibility of accidentally stepping off the cliff. Thus, the potential for future harm is no longer zero and the large penalties for stepping off the cliff may back-propagate to their antecedent states, and so on. The extent of this spread is dependent on the strength of the pessimistic belief. With increasing pessimism, the optimal policy for the agent is increasingly distance from the cliff's edge (Figure 2c). Surprisingly, at high levels of pessimistic belief, the optimal policy is to immediately step off the cliff (Figure 2d). The reason for this is simple: with high levels of pessimism, the likelihood of disaster grows large enough that any step is simply viewed as adding onto the pain.

@@

In the following sections, we present the results of simulations of pessimistic learning in a variety of MDP environments. In all simulations, the state-action values under the optimal policy, $Q_*(s,a)$, were solved for directly through Q-value iteration \citep{SuttonBarto1998, SuttonBarto1998, bertsekas2005}. The details of this algorithm are presented in the appendix. Model parameters are reported for each environment in their corresponding sections. All simulations were carried out with the python programming language and are publicly available on Github (https://github.com/szorowi1/SecretFunTimes).

>>>>

\keywords{
anxiety; avoidance; fear generalization; decision theory; computational psychiatry
}

@@

Anxiety disorders are characterized by aberrations in the cognitive processing of and response to threat. The hallmark symptom of anxiety disorders is exaggerated threat appraisal, or the tendency to perceive threat as disproportionately greater in likelihood and/or severity than is warranted. Clinical anxiety often also involves the generalization of fear, wherein an individual comes to perceive as threatening cues or situations secondarily associated with the primary fear. For example, a dog phobic individual may first fear dog parks, then parks in general, and finally any public space where a dog may plausibly be encountered. Unsurprisingly then, anxiety disorders involve persistent avoidance behavior. Avoidance behaviors in anxiety are especially pernicious as they prevent potential exposure to disconfirming evidence about perceived threats, thereby maintaining the disorder, and because they are often highly disruptive to everyday functioning. Avoidance of perceived threat often comes at considerable expense, either in foregone rewards or in the expenditure of effort. On their walk to work, the same dog phobic individual may take increasingly circuitous paths in order to avoid certain public spaces and may abandon social obligations for fear of encountering a dog. These behaviors are puzzling to theories of learning and choice behavior, not in the least because they are still observed even in safe situations posing little to no objective threat.

Some insight into anxious cognition might possibly be gleaned from a decision theoretic perspective. In decision theory, the value of potential actions (e.g. to approach or avoid) is defined in part as a function of expectations about the future. The best action is that which yields the greatest cumulative reward, contingent on current beliefs of what will likely be encountered in the future. Interestingly, anxiety disorders are associated with pessimistic beliefs about the future; anxious individuals are more likely to endorse beliefs like that they will surely face danger in the future and be unable to respond effectively to it. Despite this common finding, most studies of anxious decision making interrogate behavior in response to immediate but not future threat. Studying anxious appraisal of future threat may be more important for understanding anxiety disorders as symptoms like fear generalization and avoidance show worry for would-be but not immediate danger.

Here, we explore these issues by offering a decision theoretic account of sequential choice under pessimistic future expectations. We consider optimal behavior in deterministic, finite Markov decision processes (MDP), in which agents learn reward-maximizing (or loss-minimizing) policies in grid world environments requiring a series of choices. We first describe a fundamental asymmetry in the differential influence of reward and danger on value estimation under normative models of learning, namely that reward propagates recursively but danger does not. We then demonstrate how this asymmetry is eliminated when agents instead learn with anxious-like expectations of future action. We find that a minor deviation from normativity captures many features of anxious behavior including exaggerated threat appraisals, fear generalization, and persistent avoidance.

In the remaining sections, we also show that this simple model can account for a number of findings reported in the minority of studies that have probed anxious behavior in sequential choice environments. We then offer a novel prediction based on the the model that, consistent with negative future expectations, anxious individuals should not exhibit a free choice premium. We then close with a discussion of the relationship of our decision theoretic account to prominent clinical theories of anxiety, potential extensions to the model, and possible neural correlates.

@@

Crucially then, the value of a state under the optimal policy is contingent on an agent's particular expectations about the future. First, it is a function of the state transition distribution, or beliefs about the likelihood of encountering particular successor states conditioned on performing some action. Second, it is also a function of the actions available in those successor states, as future actions in turn define their value. Under Bellman optimality, it is normative for an agent to assume that the value of a successor state is equivalent to the best action from that state. This can be expressed as:

@@

where $q(s',a')$ denotes the a state-action value. As we will demonstrate, this assumption has profound consequences for learning about reward and danger.

In a MDP, the discovery of a rewarding action by an agent has an effect not only on the associated state but also on its antecedent states. This follows naturally from the recursive definition of state values defined above. When an agent learns a particular state predicts reward, the states which immediately precede it also acquire value (as do their antecedent states, and so on). The discovery of a rewarding action thus propagates positive value across the state space, so as to motivate the future visitation of rewarding states. A similar spread of value does not occur, however, upon the discovery of a suboptimal or harmful action. Under the optimal policy, only the best action is chosen; a reward-maximizing agent will avoid performing a harmful action insofar that another option is available. When avoidance is possible, the discovery of a harmful action will not impact the value of its associated state. Moreover, an avoidable harmful action will not influence the value of its antecedent states, as is reflected by the max operator in the equation above. Thus, the optimistic assumptions about future behavior under Bellman optimality yields a fundamental asymmetry in the treatment of reward and danger: whereas rewarding opportunity propagates recursively to antecedent states, danger does not whenever avoidance is possible.

This principle is highlight in a toy MDP, the open field environment (Figure 1a). The open field is a grid world that entirely empty except for two states, one rewarding and one harmful. Importantly, the open field MDP is deterministic, such that the one-step transition distribution respects the agent's choice. An agent in this environment then will only experience an aversive outcome if it chooses to approach the dangerous state. The state value estimates learned by an agent under Bellman optimality reflect the asymmetry described above (Figure 1b). In the open field, reward propagates recursively from the source to antecedent states whereas danger does not. Thus the agent learns to associate all states (except the harmful state itself) with positive value. Because harm is avoidable in this environment, all states (even those proximal to threat) come to represent the opportunity for reward.

What happens then when an agent believes (falsely) that danger is not avoidable? In the MDP formalism, such an errant  belief may manifest in numerous ways. First, an agent may possess inaccurate assumptions about the state transition distribution. For example, an agent may wrongly believe it possible that it could transition to a state closer to danger in spite of an avoidance action. Alternately, an agent may operate under pessimistic expectations about the availability of future actions. It may believe it has the capacity to act in accordance with the optimal (reward-maximizing) policy in the present, but may be unable to do so in the future. We note here that similar beliefs about the unreliability of self or the environment in the presence of perceived threat are often endorsed in anxiety disorders.

We can easily model pessimistic expectations about future action through a minor change to the definition of state value. Previously introduced by \cite{Gaskett2003}, we define pessimistic learning as the expectation of future deviation from the optimal policy such that the value of a successor state is equal to:

@@

Here, the parameter $w$ controls the degree of pessimism. When $\w = 1$, an agent is maximally optimistic and fully expects to act according to the best possible actions in the future. When $w = 0$, the agent is maximally pessimistic and fully expects to act according to the worst possible actions in the future. When $w = 0.5$, the weighs the best and worst possible future actions equally. We note we are not committed to this particular implementation. Pessimistic learning could be introduced through a variety of other manipulations. We have selected this particular formalism because out of convenience; it does not constitute a mechanistic claim about anxious cognition.

@@

For a pessimistic agent ($w > 0$), the expectation that future action will always be optimal (reward-maximizing) is violated; instead, the agent believes that it may act with some probability in direct contrast to its preferences. Under these conditions, an agent cannot expect to invariably avoid choosing a harmful action. Therefore, the presence of danger now poses threat and safety cannot be guaranteed. The consequences of this manipulation for the open field environment are presented in Figure 1c,d. The deviation from Bellman optimality, wherein the value of antecedent states are contingent on both the best and worst possible future actions, means that now more than reward opportunities are able to propagate recursively through state space. Indeed when $w$ is non-zero, threat value from danger also recursively propagates from its source through adjacent states. The extent of this spread is contingent on the degree of pessimistic belief. As an agent grows increasingly certain of future deleterious action (i.e. $w \rightarrow 1$), the greater the extent to which threat value permeates the state space.

We also note that this simple demonstration produces behaviors similar to the core symptoms of anxiety disorders. As is evidence in Figure 1c,d, the pessimistic agent exhibits exaggerated threat appraisals. It estimates danger at disproportionately more states than is warranted (likelihood bias) and perceives those states as more threatening than they objectively are (severity bias). Similarly, the propagation of threat value is reminiscent of the fear generalization observed in clinical anxiety. Through pessimistic learning the agent comes to regard as threatening states with increasingly distant associations to the danger source.

The model also predicts persistent avoidance behaviors similar to those observed in clinical anxiety. Importantly, the avoidance behavior exhibited by the model begins early in the chain of sequential choices. In other words, the agent performs avoidance before it is proximal to danger. This is because agents in this model act in response to anticipated future threat. Specifically, the agent cannot be certain it will not act against its preferences in the future. To minimize the likelihood of inadvertently encountering danger, the agent therefore places as much distance between itself and the possible danger through early avoidance actions. As discussed in the introduction, this is what is observed both clinically and empirically in studies of anxious individuals responding to threat \citep{Bach2014, Bach2017, Sheynin2014}. Thus our model not only predicts persistent avoidance behavior, but also captures the excessive, overly cautious profile of avoidance behavior exhibited in clinical anxiety.

@@

In the following sections, we show that our model is capable of recreating the results of several empirical studies of anxious cognition in sequential choice environments. We represent these experiments again using the formalism of MDPs. In all simulations, the values of environment states under the optimal policy were solved for directly through Q-value iteration \citep{SuttonBarto1998, SuttonBarto1998, bertsekas2005}. The details of this algorithm are presented in the appendix. All simulations were carried out with the python programming language and are publicly available on Github (https://github.com/szorowi1/SecretFunTimes).
